\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-29 papers/jfp/5-rebuilders.tex
Polish S5-S6

\subsubsection{Trace Representation}

@@

A verifying trace deliberately records only small hashes, so that it can be small.
In contrast, a \emph{constructive} trace also stores the resulting value. Concretely,
it records a list of \hs{Trace}~\hs{k}~\hs{v}.
Once we are storing the complete result it makes sense
to record many constructive traces per key, and to share them with other users,
providing cloud-build functionality. We can represent this additional
information with the operations:

\begin{minted}[fontsize=\small,xleftmargin=10pt]{haskell}
recordCT    :: k -> v -> [(k, Hash v)] -> CT k v -> CT k v
constructCT :: (Monad m, Eq k, Eq v)
            => k -> (k -> m (Hash v)) -> CT k v -> m [v]
\end{minted}

\noindent
The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing
the hash of the resulting value, we require the actual value. The \hs{verifyVT}
has been replaced with \hs{constructCT}, which instead of taking the hash of the
current value as \emph{input}, returns a list of suitable values as \emph{output}.
If the current value in the store matches one of the possible values, the build
can skip this key. If the resulting list is empty, the key must be rebuilt.
However, if the current value does not match the store, and there is a possible
value, we can use any value from the constructive list \emph{without} doing any
work to build it, and copy it into the store.

Any \hs{Applicative} build system using constructive traces, e.g.
\CloudBuild, can index directly from the key and results to the output result~--~i.e.
\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
are stored on a central server, the client can compute the key and the hashes of
its dependencies, then make a single call to the server to retrieve the result.

In practice, many cloud build systems store hashes of values in the trace information,
then have a separate content-addressable cache which associates hashes with
their actual contents.

@@

Concretely, taking the example from Figure \ref{fig-make}, to decide whether \hs{main.exe}
is out of data, a \emph{constructive} trace would look at \hs{util.o} and \hs{main.o}
(the immediate dependencies), while a \hs{deep constructive} trace would look at \hs{util.c},
\hs{util.h} and \hs{main.c}.

An advantage of deep constructive traces is that to decide if \hs{main.exe} is up to date only
requires consulting its inputs, not even considering \hs{util.o} or \hs{main.o}.
Such a feature is often known as a \emph{shallow build}, as discussed in~\S\ref{sec-cloud-aspects}.

@@

\begin{description}
\item[Deterministic tasks] If the tasks are not \emph{deterministic} then it is possible to
violate correctness, as illustrated by an example in~\S\ref{sec-cloud-aspects}
(see Fig.~\ref{fig-frankenbuild}).
\item[No early cutoff] Such traces cannot support early cutoff
(\S\ref{sec-background-shake}), since the results of intermediate computations are not considered.
\end{description}

>>>>

\subsubsection{Trace Representation}\label{sec-trace-representation}

@@

A verifying trace deliberately records only hashes, so that it can be small. In
contrast, a \emph{constructive} trace also stores the resulting value.
Concretely, it records a list of \hs{Trace}~\hs{k}~\hs{v}~\hs{v}, where
\hs{Trace} is as defined in~\S\ref{sec-trace-representation}. Once we are
storing the complete result it makes sense to record many constructive traces
per key, and to share them with other users, providing cloud-build
functionality. We can represent this additional information with the operations:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
recordCT    :: k -> v -> [(k, Hash v)] -> CT k v -> CT k v
\end{minted}
\vspace{-2mm}
\begin{minted}[xleftmargin=10pt]{haskell}
constructCT :: (Monad m, Eq k, Eq v)
            => k -> (k -> m (Hash v)) -> CT k v -> m [v]
\end{minted}
\vspace{1mm}

\noindent
The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing
the hash of the resulting value, we require the actual value. The \hs{verifyVT}
has been replaced with \hs{constructCT}, which instead of taking the hash of the
current value as \emph{input}, returns a list of possible values as
\emph{output} -- there may be more than one, because some build tools are
non-deterministic, see~\S\ref{sec-non-determinism}.

Regardless of the chosen scheduler~(\S\ref{sec-scheduler}), there are three
cases to consider when using a rebuilder based on constructive traces:

\begin{itemize}
  \item If \hs{constructCT} returns the empty list of possible values, the key
        must be rebuilt.
  \item If the current value in the store matches one of the possible values,
        the build system can skip this key. Here a constructive trace is used
        for verifying the current value.
  \item If the current value in the store does not match any possible value, we
        can use any of the possible values \emph{without} doing any work to
        build it, and copy it into the store.
\end{itemize}

Any \hs{Applicative} build system using constructive traces, e.g.
\CloudBuild~\S\ref{sec-implementation-cloud}, can index directly from the
key and the hashes of its dependencies to the output result~--~i.e.
\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
are stored on a central server, the client can compute the key and the hashes of
its dependencies locally, and then make a single call to the server to retrieve
the result.

In practice, many cloud build systems store hashes of values in the trace store,
i.e. use \hs{Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v)} trace entries just like the
verifying traces, and have a separate content-addressable cache which associates
hashes with their actual contents.

@@

Concretely, taking the example from Fig.~\ref{fig-make}, to decide whether
\cmd{main.exe} is out of date, a \emph{constructive} trace would look at
\cmd{util.o} and \cmd{main.o} (the immediate dependencies), whereas a
\emph{deep constructive} trace would look at \cmd{util.c}, \cmd{util.h} and
\cmd{main.c}.

An advantage of deep constructive traces is that to decide if \cmd{main.exe} is
up to date only requires consulting its inputs, not even considering
\cmd{util.o} or \cmd{main.o}. Such a feature is often known as a
\emph{shallow build}, as discussed in~\S\ref{sec-cloud-aspects}.

@@

\begin{itemize}
  \item \textbf{Tasks must be deterministic}: If the tasks are not
        \emph{deterministic} then it is possible to violate correctness, as
        illustrated by an example in~\S\ref{sec-cloud-aspects}, see
        Fig.~\ref{fig-frankenbuild}.
  \item \textbf{No early cutoff}: Deep constructive traces cannot support early
        cutoff (\S\ref{sec-background-shake}), since the results of intermediate
        computations are not considered.
\end{itemize}

### 2019-04-29 papers/jfp/6-implementations.tex
Polish S5-S6

\section{Build systems, concretely}\label{sec-implementations} \label{sec-design-space}

\begin{table}
\caption{Build systems \`a la carte.\label{tab-build-systems}}
\vspace{-4mm}
% \smaller
\centering
\begin{tabular}{lccc}
\hline
 & \multicolumn{3}{c}{Scheduling algorithm} \\
Rebuilding strategy\gap  & \gap{}Topological\gap\S\ref{sec-topological}\gap & \gap{}Restarting\gap\S\ref{sec-restarting}\gap & \gap{}Suspending\gap\S\ref{sec-suspending}\gap    \\\hline
Dirty bit\hfill\S\ref{sec-dirty-bit}                                                             & \Make       & \Excel & -              \\
Verifying traces\hfill\S\ref{sec-verifying-traces}                                               & \Ninja      & -      & \Shake         \\
Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & -              \\
Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline
\end{tabular}
\vspace{-4mm}
\end{table}

@@

\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v =
    k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
\vspace{0.5mm}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system; stores current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTimes) <- get
    let dirty = case Map.lookup key modTimes of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTimes > time) (dependencies task)
    if not dirty then return value else do
        put (now + 1, Map.insert key now modTimes)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Scheduler Applicative i i k v
topological rebuilder tasks target = execState $ mapM_ build order
  where
    build :: k -> State (Store i k v) ()
    build key = case tasks key of
        Nothing -> return ()
        Just task -> do
            store <- get
            let value = getValue key store
                newTask :: Task (MonadState i) k v
                newTask = rebuilder key value task
                fetch :: k -> State i v
                fetch k = return (getValue k store)
            newValue <- liftStore (run newTask fetch)
            modify $ putValue key newValue
    order = topSort (reachable dep target)
    dep k = case tasks k of { Nothing -> []; Just task -> dependencies task }
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> [k] -- Throws error on a cyclic graph
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation
liftStore :: State i a -> State (Store i k v) a
liftStore x = do
    (a, newInfo) <- gets (runState x . getInfo)
    modify (putInfo newInfo)
    return a
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-5mm}
\end{figure}

@@

The rebuilder receives three arguments: a \hs{key}, its current \hs{value}, and
the applicative \hs{task} that can be used to rebuild the \hs{key} if necessary.
The rebuilder first decides if the \hs{key} is \hs{dirty} by consulting
\hs{modTimes}: if the \hs{key} is not found, that must mean it has never been
built before; otherwise \hs{modTimeRebuilder} can see if any of the \hs{task}'s
dependencies (computed by \hs{dependencies}) are out of date. If the \hs{key} is
\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
the new modification time of the \hs{key}\footnote{The real \Make
relies on the file system to track file modification times, but we prefer
to make it explicit here.}; otherwise we can just return the current \hs{value}.

@@

The pre-processing stage uses the function \hs{dependencies}, defined
in~\S\ref{sec-deps}, to extract static dependencies from the provided
applicative \hs{task}. We compute the linear processing \hs{order} by
constructing the graph of keys \hs{reachable} from the \hs{target} via
dependencies, and performing the topological sort of the result. We omit
implementation of textbook graph algorithms \hs{reachable} and \hs{topSort},
e.g. see~\cite{cormen2001introduction}.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system; stores a dirty bit per key and calc chain
type Chain k = [k]
type ExcelInfo k = (k -> Bool, Chain k)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting dirtyBitRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on dirty bits
dirtyBitRebuilder :: Rebuilder Monad (k -> Bool) k v
dirtyBitRebuilder key value task = Task $ \fetch -> do
    isDirty <- get
    if isDirty key then run task fetch else return value
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Scheduler Monad (ir, Chain k) ir k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- liftChain $ go Set.empty $ chain ++ [target | target `notElem` chain]
    modify $ mapInfo $ \(ir, _) -> (ir, newChain)
  where
    go :: Set k -> Chain k -> State (Store ir k v) (Chain k)
    go _    []       = return []
    go done (key:keys) = case tasks key of
        Nothing -> (key :) <$> go (Set.insert key done) keys
        Just task -> do
            store <- get
            let newTask :: Task (MonadState ir) k (Either k v)
                newTask = try $ rebuilder key (getValue key store) task
                fetch :: k -> State ir (Either k v)
                fetch k | k `Set.member` done = return $ Right (getValue k store)
                        | otherwise           = return $ Left k
            result <- liftStore (run newTask fetch) -- liftStore is defined in Fig. 7
            case result of
                Left dep -> go done $ dep : filter (/= dep) keys ++ [key]
                Right newValue -> do modify $ putValue key newValue
                                     (key :) <$> go (Set.insert key done) keys
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task that accepts a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation (implementation omitted)
liftChain :: State (Store ir k v) a -> State (Store (ir, Chain [k]) k v) a
\end{minted}
\vspace{-2.5mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-2.5mm}
\end{figure}

@@

Our model of \Shake is shown in Fig.~\ref{fig-shake-implementation}. It stores
verifying traces \hs{VT}~\hs{k}~\hs{v} defined in~\S\ref{sec-verifying-traces}
as persistent build information and is composed of the \hs{suspending} scheduler
and the \hs{vtRebuilder}.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system; stores verifying traces
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    upToDate <- verifyVT key (hash value) (fmap hash . fetch) =<< get
    if upToDate then return value else do
        (newValue, deps) <- track task fetch
        modify $ recordVT key (hash newValue) [ (k, hash v) | (k, v) <- deps ]
        return newValue
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Scheduler Monad i i k v
suspending rebuilder tasks target store = fst $ execState@\,@(fetch target)@\,@(store,@\,@Set.empty)
  where
    fetch :: k -> State (Store i k v, Set k) v
    fetch key = do
        done <- gets snd
        case tasks key of
            Just task | key `Set.notMember` done -> do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                newValue <- liftRun newTask fetch
                modify $ \(s, d) -> (putValue key newValue s, Set.insert key d)
                return newValue
            _ -> gets (getValue key . fst) -- fetch the existing value
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Run a task using a callback that operates on a larger state (implementation omitted)
liftRun :: Task (MonadState i) k v
        -> (k -> State (Store i k v, Set k) v) -> State (Store i k v, Set k) v
\end{minted}
\vspace{-1mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
\vspace{-3mm}
\end{figure}

The rebuilder executes the verification query \hs{verifyVT} to determine if the
\hs{key} is \hs{upToDate}. If it is, the rebuilder simply returns the \hs{key}'s
current \hs{value}. Otherwise it executes the \hs{task}, obtaining both a
\hs{newValue} and the \hs{key}'s dynamic dependencies \hs{deps} (see the
definition of \hs{track} in~\S\ref{sec-deps}), which are subsequently recorded
in a new verification trace using \hs{recordVT}.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Bazel build system; stores constructive traces
bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
bazel = restarting2 ctRebuilder -- implementation of 'restarting2' is omitted (22 lines)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A rebuilder based on constructive traces
ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
ctRebuilder key value task = Task $ \fetch -> do
    cachedValues <- constructCT key (fmap hash . fetch) =<< get
    case cachedValues of
        _ | value `elem` cachedValues -> return value
        cachedValue:_ -> return cachedValue
        [] -> do (newValue, deps) <- track task fetch
                modify $ recordCT key newValue [ (k, hash v) | (k, v) <- deps ]
                return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 9
cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
cloudShake = suspending ctRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- CloudBuild build system, implementation of 'topological' is given in Fig. 7
cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
cloudBuild = topological (adaptRebuilder ctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a monadic rebuilder to the corresponding applicative one
adaptRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
adaptRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Buck build system, implementation of 'topological' is given in Fig. 7
buck :: (Ord k, Hashable v) => Build Applicative (DCT k v) k v
buck = topological (adaptRebuilder dctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on deep constructive traces, analogous to 'ctRebuilder'
dctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (DCT k v) k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Nix build system, implementation of 'suspending' is given in Fig. 9
nix :: (Ord k, Hashable v) => Build Monad (DCT k v) k v
nix = suspending dctRebuilder
\end{minted}
\vspace{-2mm}
\caption{\Bazel, \Cloud \Shake, \CloudBuild, \Buck and \Nix implemented in
our framework; \hs{restarting2} differs from \hs{restarting} used by \Excel
(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the calc
chain.}
\label{fig-cloud-implementations}
\vspace{-2mm}
\end{figure}

@@

@@

Our models of \Buck~\cite{buck} and \Nix~\cite{dolstra2004nix} use the
\hs{dctRebuilder}, a rebuilder based on deep constructive traces
(\S\ref{sec-deep-constructive-traces}), whose implementation we omit since it
is very similar to that of \hs{ctRebuilder}. \Buck uses the \hs{topological}
scheduler and is therefore an applicative build system, whereas \Nix uses the
\hs{suspending} scheduler and is monadic.

>>>>

\section{Build systems, concretely}\label{sec-implementations}\label{sec-design-space}

\begin{table}
\caption{Build systems \`a la carte.\label{tab-build-systems}}
\vspace{-2mm}
% \smaller
\centering
\begin{tabular}{lccc}
\hline
 & \multicolumn{3}{c}{\textbf{Scheduling algorithm}}\vspace{1.5mm} \\
\textbf{Rebuilding strategy}\gap  & \gap{}Topological\gap\S\ref{sec-topological}\gap & \gap{}Restarting\gap\S\ref{sec-restarting}\gap & \gap{}Suspending\gap\S\ref{sec-suspending}\gap    \\\hline
Dirty bit\hfill\S\ref{sec-dirty-bit}                                           & \Make       & \Excel & -              \\
Verifying traces\hfill\S\ref{sec-verifying-traces}                             & \Ninja      & -      & \Shake         \\
Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}           & \CloudBuild & \Bazel & -              \\
Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline
\end{tabular}
\vspace{-2mm}
\end{table}

@@

\vspace{1mm}
\begin{minted}[fontsize=\small,xleftmargin=5pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task@\,@(MonadState ir) k v
\end{minted}
\vspace{1mm}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system; stores current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTimes) <- get
    let dirty = case Map.lookup key modTimes of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTimes > time) (dependencies task)
    if not dirty then return value else do
        put (now + 1, Map.insert key now modTimes)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Scheduler Applicative i i k v
topological rebuilder tasks target = execState $ mapM_ build order
  where
    build :: k -> State (Store i k v) ()
    build key = case tasks key of
        Nothing -> return ()
        Just task -> do
            store <- get
            let value = getValue key store
                newTask :: Task (MonadState i) k v
                newTask = rebuilder key value task
                fetch :: k -> State i v
                fetch k = return (getValue k store)
            newValue <- liftStore (run newTask fetch)
            modify $ putValue key newValue
    order = topSort (reachable dep target)
    dep k = case tasks k of { Nothing -> []; Just task -> dependencies task }
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> [k] -- Throws error on a cyclic graph
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation
liftStore :: State i a -> State (Store i k v) a
liftStore x = do
    (a, newInfo) <- gets (runState x . getInfo)
    modify (putInfo newInfo)
    return a
\end{minted}
% \vspace{-3mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
% \vspace{-5mm}
\end{figure}

@@

The rebuilder receives three arguments: a \hs{key}, its current \hs{value}, and
the applicative \hs{task} that can be used to rebuild the \hs{key} if necessary.
The rebuilder first decides if the \hs{key} is \hs{dirty} by consulting
\hs{modTimes}: if the \hs{key} is not found, that must mean it has never been
built before; otherwise \hs{modTimeRebuilder} can see if any of the \hs{task}'s
dependencies (computed by \hs{dependencies}) are out of date. If the \hs{key} is
\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
the new modification time of the \hs{key}\footnote{The real \Make relies on the
file system to track file modification times, but we prefer to make this
explicit in our model.}; otherwise we can just return the current \hs{value}.

@@

The pre-processing stage uses the function \hs{dependencies}, defined
in~\S\ref{sec-deps}, to extract static dependencies from the provided
applicative \hs{task}. We compute the linear processing \hs{order} by
constructing the graph of keys \hs{reachable} from the \hs{target} via
dependencies, and performing the topological sort of the result. We omit
implementation of textbook graph algorithms \hs{reachable} and \hs{topSort},
e.g. see~Cormen~\etal~\shortcite{cormen2001introduction}.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system; stores a dirty bit per key and calc chain
type Chain k = [k]
type ExcelInfo k = (k -> Bool, Chain k)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting dirtyBitRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on dirty bits
dirtyBitRebuilder :: Rebuilder Monad (k -> Bool) k v
dirtyBitRebuilder key value task = Task $ \fetch -> do
    isDirty <- get
    if isDirty key then run task fetch else return value
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Scheduler Monad (ir, Chain k) ir k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- liftChain $ go Set.empty
                          $ chain ++ [target | target `notElem` chain]
    modify $ mapInfo $ \(ir, _) -> (ir, newChain)
  where
    go :: Set k -> Chain k -> State (Store ir k v) (Chain k)
    go _    []       = return []
    go done (key:keys) = case tasks key of
      Nothing -> (key :) <$> go (Set.insert key done) keys
      Just task -> do
        store <- get
        let newTask :: Task (MonadState ir) k (Either k v)
            newTask = try $ rebuilder key (getValue key store) task
            fetch :: k -> State ir (Either k v)
            fetch k | k `Set.member` done = return $ Right (getValue k store)
                    | otherwise           = return $ Left k
        result <- liftStore (run newTask fetch) -- liftStore is in Fig. 8
        case result of
            Left dep -> go done $ dep : filter (/= dep) keys ++ [key]
            Right newValue -> do modify $ putValue key newValue
                                 (key :) <$> go (Set.insert key done) keys
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task that accepts a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation (omitted)
liftChain :: State (Store ir k v) a -> State (Store (ir, Chain [k]) k v) a
\end{minted}
% \vspace{-2.5mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
% \vspace{-2.5mm}
\end{figure}

@@

Our model of \Shake (Fig.~\ref{fig-shake-implementation}) stores verifying
traces \hs{VT}~\hs{k}~\hs{v} defined in~\S\ref{sec-verifying-traces} as
persistent build information and is composed of the \hs{suspending} scheduler
and the \hs{vtRebuilder}.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system; stores verifying traces
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    upToDate <- verifyVT key (hash value) (fmap hash . fetch) =<< get
    if upToDate then return value else do
        (newValue, deps) <- track task fetch
        modify $ recordVT key (hash newValue) [ (k, hash v) | (k, v) <- deps ]
        return newValue
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Scheduler Monad i i k v
suspending rebuilder tasks target store =
    fst $ execState (fetch target) (store, Set.empty)
  where
    fetch :: k -> State (Store i k v, Set k) v
    fetch key = do
        done <- gets snd
        case tasks key of
            Just task | key `Set.notMember` done -> do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                newValue <- liftRun newTask fetch
                modify $ \(s, d) -> (putValue key newValue s, Set.insert key d)
                return newValue
            _ -> gets (getValue key . fst) -- fetch the existing value
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Run a task using a callback that operates on a larger state (omitted)
liftRun :: Task (MonadState i) k v
        -> (k -> State (Store i k v, Set k) v) -> State (Store i k v, Set k) v
\end{minted}
% \vspace{-1mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
% \vspace{-3mm}
\end{figure}

The rebuilder performs the \hs{verifyVT} query to determine if the \hs{key} is
\hs{upToDate}. If it is, the rebuilder simply returns the \hs{key}'s current
\hs{value}. Otherwise it executes the \hs{task}, obtaining both a \hs{newValue}
and the \hs{key}'s dynamic dependencies \hs{deps} (see the definition of
\hs{track} in~\S\ref{sec-deps}), which are subsequently recorded in the trace
store using \hs{recordVT}.

@@

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Bazel build system; stores constructive traces
bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
bazel = restartingQ ctRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting scheduler based on a build queue, omitted (22 lines)
restartingQ :: (Hashable v, Eq k) => Scheduler Monad (CT k v) (CT k v) k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A rebuilder based on constructive traces
ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
ctRebuilder key value task = Task $ \fetch -> do
    cachedValues <- constructCT key (fmap hash . fetch) =<< get
    case cachedValues of
        _ | value `elem` cachedValues -> return value
        cachedValue:_ -> return cachedValue
        [] -> do (newValue, deps) <- track task fetch
                modify $ recordCT key newValue [ (k, hash v) | (k, v) <- deps ]
                return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 10
cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
cloudShake = suspending ctRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- CloudBuild build system, implementation of 'topological' is given in Fig. 8
cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
cloudBuild = topological (adaptRebuilder ctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a monadic rebuilder to the corresponding applicative one
adaptRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
adaptRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Buck build system, implementation of 'topological' is given in Fig. 8
buck :: (Ord k, Hashable v) => Build Applicative (DCT k v) k v
buck = topological (adaptRebuilder dctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on deep constructive traces, analogous to 'ctRebuilder'
dctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (DCT k v) k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Nix build system, implementation of 'suspending' is given in Fig. 10
nix :: (Ord k, Hashable v) => Build Monad (DCT k v) k v
nix = suspending dctRebuilder
\end{minted}
% \vspace{-2mm}
\caption{\Bazel, \Cloud \Shake, \CloudBuild, \Buck and \Nix in our framework.}
\label{fig-cloud-implementations}
% \vspace{-2mm}
\end{figure}

@@

Our models of \Buck~\cite{buck} and \Nix~\cite{dolstra2004nix} use the rebuilder
based on deep constructive traces (\S\ref{sec-deep-constructive-traces}), called
\hs{dctRebuilder}, whose implementation we omit since it is very similar to that
of \hs{ctRebuilder}. \Buck uses the \hs{topological} scheduler and is therefore
an applicative build system, whereas \Nix uses the \hs{suspending} scheduler and
is monadic.

\todo{AM}{Add implementation of \hs{dctRebuilder}?}
